---
title: "Homework Assignment 2"
author: "DS4043, Spring 2023"
date: "__Due on March 29, 2023 at 11:59 pm__"
output: pdf_document
---


```{r setup, include=FALSE}
library(knitr)
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
#library(tidyverse)
#library(reshape2)
```


1. Consider the multivariate normal distribution vector $\mathrm{X}=\left(\mathrm{X}_{1}, \mathrm{X}_{2}, \mathrm{X}_{3}\right)^{\mathrm{T}}$ having mean vector $\boldsymbol{\mu}=(0,1,2,)^{\mathrm{T}}$ and covariance matrix

$$
  \boldsymbol{\Sigma}=\left[\begin{array}{ccc}
                            1 & -0.5 & 0.5 \\
                            -0.5 & 1 & -0.5 \\
                            0.5 & -0.5 & 1
                            \end{array}\right]
$$
    
a) Generate 100 random observations from the multivariate normal distribution given above with _set.seed(12)_. (Hint: see _?mvrnorm_ ) You may need to use the package MASS.

```{r indent=indent1}
# install.packages("MASS")
library(MASS) # you may need to use this package 
set.seed(12)

miu <- c(0, 1, 2)
Sigma <- matrix(c(1, -0.5, 0.5, -0.5, 1, -0.5, 0.5, -0.5, 1), ncol = 3)
X <- mvrnorm(n = 100, mu = miu, Sigma = Sigma)
head(X)
```

b) Construct a scatterplot matrix for $\mathbf{X}$ and add a fitted smooth density curve on the diagonal panels for each $\mathrm{X}_{1}, \mathrm{X}_{2}, \mathrm{X}_{3}$ to verify that the location and correlation for each plot agrees with the parameters of the corresponding bivariate distributions.

```{r}
# install.packages("car")
#library(car)
#scaterplotMatrix(X, main = 'Scatterplot Matrix')
# install.packages("psych")
library(psych)
pairs.panels(X, 
             smooth = TRUE,
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
# method = "pearson", # correlation method
```

c) Obtain the correlation plot for the generated sample $\mathbf{X}$, where coefficients are added to the plot whose magnitude are presented by different colors. Let the visualization method of correlation matrix to be ellipse.

```{r indent=indent1,message=FALSE}
# install.packages("corrplot")
library(corrplot)  # you may need to use this package 

M <- cor(X)
corrplot(M, method = "ellipse", type = "lower", tl.col = "black", tl.srt = 45)
```

    
d) Given the covariance matrix $\Sigma$, find $\sigma_{x_1}$, $\sigma_{x_2}$ and $\rho_{x_1x_2}$. Consider the joint PDF of bivariate normal distribution

$$
  \begin{aligned}
&f_{X Y}(x, y)=\frac{1}{2 \pi \sigma_{X} \sigma_{Y} \sqrt{1-\rho^{2}}} \\
&\exp \left\{-\frac{1}{2\left(1-\rho^{2}\right)}\left[\left(\frac{x-\mu_{X}}{\sigma_{X}}\right)^{2}+\left(\frac{y-\mu_{Y}}{\sigma_{Y}}\right)^{2}-2 \rho \frac{\left(x-\mu_{X}\right)\left(y-\mu_{Y}\right)}{\sigma_{X} \sigma_{Y}}\right]\right\},
\end{aligned}
$$
sketch a surface plot for $X_{1}$ and  $X_{2}$, based on their bivariate probability density function. (Hint: if you want to use _curve3d_, please install and use the package _emdbook_)

```{r indent=indent1, warning=FALSE}
# install.packages("emdbook")
library(emdbook) # you may need to use this package 

# calculate mean of x.
mean_x1 <- mean(X[, 1])
mean_x2 <- mean(X[, 2])
# calculate the standard derivation of x.
covX <- cov(X)
sigma_x1 <- sqrt(covX[1, 1])
sigma_x2 <- sqrt(covX[2, 2])
rho_x1x2 <- covX[1, 2] / (sigma_x1 * sigma_x2)

f <- function(x, y) {
  term1 <- ((X[, 1]-mean_x1)/sigma_x1)^2
  term2 <- ((X[, 2]-mean_x2)/sigma_x2)^2
  term3 <- -2*rho_x1x2*(X[, 1]-mean_x1)*(X[, 2]-mean_x2)/(sigma_x1 * sigma_x2)
  z <- 1/(2 * pi * sigma_x1 * sigma_x2 * sqrt(1-rho_x1x2^2)) * exp(-1/(2*(1-rho_x1x2^2))*(term1+term2+term3))
  dmvnorm(cbind(x, y), mu = c(mean_x1, mean_x2), Sigma = covX[1:2, 1:2])
}

curve3d(f(x, y), xlim = c(-3, 3), ylim = c(-3, 3), sys3d = "persp")
```
  
e) Sketch 3-D scatter plots for each of $X_{1}, X_{2}$ and $X_{3}$ as a $z$ axis and rest two variables as $x$ and $y$ axes. Put these 3 plots in one picture.

```{r indent=indent1, warning=FALSE}
library(lattice)

trellis.par.set(layout.heights = list(top.padding = 0, bottom.padding = 0))
trellis.par.set(layout.widths = list(left.padding = 0, right.padding = 0))

p1 <- cloud(X[, 1] ~ X[, 2] * X[, 3], screen = list(z = -60, x = -60), xlab = "X2", ylab = "X3", zlab = "X1", par.settings = list(layout.widths = list(left.padding = 0)))
p2 <- cloud(X[, 2] ~ X[, 1] * X[, 3], screen = list(z = -60, x = -60), xlab = "X1", ylab = "X3", zlab = "X2")
p3 <- cloud(X[, 3] ~ X[, 1] * X[, 2], screen = list(z = -60, x = -60), xlab = "X1", ylab = "X2", zlab = "X3")

print(p1, split = c(1, 1, 3, 1), more = TRUE)
print(p2, split = c(2, 1, 3, 1), more = TRUE)
print(p3, split = c(3, 1, 3, 1))
```


2. A continuous random variable $X$ has the probability density function \[f_X(t)=\begin{cases} at+bt^2 & 0<t<1\\ 0 & \text{otherwise}
\end{cases}.\] If $E[X]=1/2$, find
(a) $a$ and $b$
$$\int(f_X(t))dt = \int(at+bt^2)dt\\
     \Rightarrow 1/2at_2 + 1/3bt^3\\
    \Rightarrow 1/2a + 1/3b = 1$$

$$ E(X) = E(f_X(t)) = \int(t*f_X(t))dt\\
     \Rightarrow \int(at^2 + bt^3)dt\\
     \Rightarrow 1/3at^3 + 1/4bt^4\\
     \Rightarrow 1/3a + 1/4b = 1/2$$

$$ 1/2a + 1/3b = 1\\
 1/3a + 1/4b = 1/2\\
\therefore a = 6, b = -6$$
(b) $P(X<1/2)$
$$ P(X<1/2) = \int(6t-6t^2)dt = 3t^2-2t^3 = 0.5$$

(c) $\operatorname{Var}(X)$
$$ E(X^2) = \int(t^2(6t-6t^2))dt = 3/2t^4-6/5t^5 = 3/10\\
 \because Var(X) = E(X^2)-(E(X))^2\\
 \therefore Var(X) = 3/10 - 1/4 = 1/20$$

(d) Generate the density plot of $X$

```{r indent=indent1,message=FALSE}
# !!! (a), (b), (c) are on paper.
# (a) a = 6; b = -6
# (b) P(X<1/2) = 0.5
# (c) Var(X) = 1/20
# (d) Generate the density plot of X.
X <- runif(100)
d <- curve(6*x - 6*x^2, 0, 1, main = 'density funcion')
#d <- curve(3*x^2-2*x^3, 0, 1)
d
```
	

3. Consider a nonparametric regression model
$$
y_{i}=g\left(x_{i}\right)+\epsilon_{i}, \quad 1 \leq i \leq n,
$$
where $y_{i}$'s are observations, $g$ is an unknown function, and $\epsilon_{i}$'s are independent and identically distributed random errors with zero mean and variance $\sigma^{2}$. $n$ is the number of observations. Usually one fits the mean function $g$ first and then estimates the variance $\sigma^{2}$ from residual sum of squares $\hat{\sigma}^{2} = \sum_{i=1}^{n} \hat{\epsilon}_i^2/(n-1)$ where $\hat{\epsilon}_i = y_i - \hat{g}(x_i)$. However this method requires an estimate of the unknown function $g$. 
Then some researchers proposed some difference-based estimators which does not require the estimation of $g$. Assume that $x$ is univariate and $0 \leq x_{1} \leq \cdots \leq x_{n} \leq 1 .$ Rice (1984) proposed the first order difference-based estimator
$$
\hat{\sigma}_{R}^{2}=\frac{1}{2(n-1)} \sum_{i=2}^{n}\left(y_{i}-y_{i-1}\right)^{2}.
$$
Gasser, Sroka and Jennen-Steinmetz (1986) proposed the second order difference based estimator and for equidistant design points (i.e. $x_i$ and $x_{i+1}$ have the same distance for all $i=1,2,\dots, n$), $\hat{\sigma}_{G S J}^{2}$ reduces to
$$
\hat{\sigma}_{G S J}^{2}=\frac{2}{3(n-2)} \sum_{i=2}^{n-1}\left(\frac{1}{2} y_{i-1}-y_{i}+\frac{1}{2} y_{i+1}\right)^{2} .
$$    
Consider the temperature anomaly dataset. Temperature anomalies in degrees Celsius are based on the new version HadCRUT4 land-sea dataset (Morice et al., 2012). We focus on the global median annual temperature anomalies from 1850 to 2019 relative to the 1961-1990 average. 
We try to build up the model between time and 
global median temperature $y_i$ and year $x_i$.  
    
(a) Use _read.csv_ to read the temperature anomaly dataset. Let $x$ be the vector of years from 1850-2019, $y$ be the vector of corresponding global median annual temperature anomalies, and $n$ be the number of observations

```{r indent=indent1,message=FALSE}
data <- read.csv('temperature-anomaly.csv')
#x <- data[data[,'Entity'] == 'Global', 'Year']
x <- data[data$Entity == 'Global', 'Year']
y <- data[data$Entity == 'Global', 'Median']
n <- length(x)
# show the results.
head(x)
head(y)
print(n)
```

(b) Display a scatter plot between global median annual temperature anomalies and years with caption "Global median land-sea temperature anomaly relative to the 1961-1990 average temperature", $x-$label years and $y-$label temperature anomalies.

```{r indent=indent1,message=FALSE}
plot(x, y, xlab = 'years', ylab = 'temperature anomalies', main = 'Global median land-sea temperature anomaly relative to the 1961-1990 average temperature')
```

(c) Change the years $x$ to a new vector $x$ such that $x_i = i/n$. Compute the first order difference-based estimator. (Note: the change of $x$ or not will not affect the computation of the estimator)

```{r indent=indent1,message=FALSE}
i <- 1:n
xi <- i / n
#xi <- seq_along(x)/n
estimator_1st <- function(y){
  1/(2*(n-1)) * sum(diff(y)^2)
}
y_est_1st = estimator_1st(y)
y_est_1st
```

(d) Compute the second order difference-based estimator.

```{r indent=indent1,message=FALSE}
estimator_2nd <- function(y){
  2 * sum((0.5*y[3:n] - y[2:(n-1)] + 0.5*y[1:(n-2)])^2)/(3*(n-2))
}
y_est_2nd <- estimator_2nd(y)
y_est_2nd
```

    
 