---
title: "Homework Assignment 1"
author: "DS4043, Spring 2022"
date: "__Due on Feb 27, 2022 at 11:59 pm__"
output: pdf_document
---


```{r setup, include=FALSE}
library(knitr)
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
#library(tidyverse)
#library(reshape2)
```


1. a) Generate a random sample $X_1, ., X_{100}$ which is from a normal distribution with mean $\mu = 5$ and standard deviation $\sigma = 3$. Use set.seed(99) for random number generation.

        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
set.seed(99)
x <- rnorm(100,5,3)
x
```
        <!-- solution end -->
        
    b) Write a R function 'fx' in R to implement the function $y=(x-a)/b$, which will transform an input vector $x$ and return the output $y$. However, the function should take three input arguments $x, a$ and $b$.

        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
fx <- function(x,a,b){
  y <- (x-a)/b
  return(y)
}
```
        <!-- solution end -->
        
    c) Generate the random sample $y$ using the function in b), where $x$=the random sample generated in a), a=5 and b=3. What is the distribution of $y$? And explain. Then calculate the sample mean and standard deviation of $y$ and compare them with the population mean and standard deviation. Note, please show your derivation of the population mean and standard deviation. 

        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
y<- fx(x,a=5,b=3);
mean(y);
sd(y)
```
We know that $y = (x-a)/b$ and $x$ is following a normal distribution. Then $y$ is following normal distribution because $y$ is a linear combination of $x$. $E(y) = (E(x) - a)/b = (5-5)/3=0$ and $Var(y) = Var(x)/b^2 = 1$. So the sample mean and standard deviation of $y$ is -0.104 and 0.9007. 
        <!-- solution end -->
        
    d) Display a probability histogram of the random sample $y$ and add an estimated probability density function to your histogram 
    

        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
hist(y, freq=FALSE, main="Density histogram of y")
lines(density(y))
```
        <!-- solution end -->

    e) Add the true probability density function to your histogram in d)
    
        <!-- solution begin -->
        __Solution__:
```{r indent=indent2, cache=TRUE}
xfit<-seq(min(y),max(y),length=40)
yfit<-dnorm(xfit,0,sd=1)
hist(y, freq=FALSE, main="Density histogram of y")
lines(density(y))
lines(xfit, yfit, col="blue", lwd=2)
```
        <!-- solution end -->



2. We will use the dataset called `hflights`. This dataset contains all flights departing from Houston airports IAH (George Bush Intercontinental) and HOU (Houston Hobby). 
The data comes from the Research and Innovation Technology Administration at the Bureau of Transportation statistics: [hflights](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0). Make sure you have installed the packages `hflights` before using them.

```{r start, indent=indent1}
# Load packages
# install.packages("hflights")
library(hflights)
data(hflights)
```
    
    (a) How many rows and columns of `hflights`? Get the names of the columns. (show your codes and results)
    
        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
nrow(hflights);
ncol(hflights);
names(hflights)
```
        <!-- solution end --> 
        
    (b) Select the first 15 rows make it a data frame called `phflights`. Suppose we would like check three variables, DepTime, ArrTime and FlightNum. Select these three columns and call it `sflights`. Show the first few lines of `sflights`.
    
        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
phflights <- hflights[1:15,];
phflights
sflights <- hflights[, c("DepTime", "ArrTime", "FlightNum")]
head(sflights)
```
        <!-- solution end --> 
    
    (c) Create a new column vector Called `BNum` indicating if the FlightNum is greater than 1000 and append this column to `sflights`. Show the first few lines.
    
        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
BNum <- hflights$FlightNum > 1000
sflights <- data.frame(sflights, BNum)
head(sflights)
```
        <!-- solution end -->
        
    (d) Compute the average arrival delay (ArrDelay) to each destination for `hflights`. (Hint: use na.rm = TRUE to remove missing values) Show the first few results. Then for each carrier, calculate the percentage of flights cancelled or diverted.
    
        <!-- solution begin -->
        __Solution__:
```{r indent=indent2}
head(by(hflights$ArrDelay, hflights$Dest, mean, na.rm = TRUE));
by(hflights$Cancelled+hflights$Diverted,hflights$UniqueCarrier , mean, na.rm = TRUE)
```
        <!-- solution end -->     
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
    
