---
title: "DS4043 - MIDTERM Examination"
author: 'Name: Ruoxin WANG         Grade: _________'
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
library(knitr)
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
#library(tidyverse)
#library(reshape2)
```
## Coding Part

3. In this question, we will use the dataset called _painters_. We can simply call the data

```{r indent=indent1}
library(MASS)
data(painters)
```

(a) Use help function to find the description of this data. 

    The description is: The subjective assessment, on a 0 to 20 integer scale, of 54 classical painters. The painters were assessed on four characteristics: composition, drawing, colour and expression. The data is due to the Eighteenth century art critic, de Piles.

(b) Select the data such that the _Composition_ score is greater or equal to 15 and make it a data frame called _HCpainters_. Print out _HCpainters_.

```{r indent=indent1}
# Put your code here
HCpainters <- data.frame(painters[painters$Composition >= 15,])
HCpainters
```
    
(c) Create a new column vector Called _LowColour_ indicating if the _Colour_ score is less than 11 and append this column to _painters_. Show the first 10 rows. (This question is not related to (b))  
```{r indent=indent1}
# Put your code here
LowColour <- painters$Colour < 11
painters <- data.frame(painters, LowColour)
head(painters, 10)
``` 

(d) Obtain the scatterplot matrix of the data _painters_ for the _Composition_, _Colour_ and _Expression_. Which pair of variables have the strongest positive linear relation.

```{r indent=indent1, warning=FALSE}
# Put your code here
panel.d <- function(x, ...) {
	usr <- par("usr"); on.exit(par(usr));
	par(usr = c(usr[1:2], 0, .5))
	lines(density(x))  # add density curve.
}
x <- scale(painters[, c('Composition', 'Colour', 'Expression')])
r <- range(x)
pairs(x, diag.panel = panel.d, xlim = r, ylim = r)
```      
Solution (Which pair): Composition and Expression.

(e) Find the correlation matrix of _Composition_, _Drawing_, _Colour_ and _Expression_. Then get the correlation plot and make the visualization method as ellipse, the text label as red color, and use "diag=FALSE".

```{r message=FALSE, indent=indent1, out.width="50%"}
library(corrplot)
# Put your code here
corrMat <- cor(painters[, 1:4]);
corrplot(corrMat, type = "upper", method = "ellipse", addCoef.col = "black", diag=FALSE, tl.col="red", tl.srt=45);
```

(f) Plot the Boxplot of _Composition_ by groups of _School_.

```{r indent=indent1, out.width="50%"}
# Put your code here
boxplot(painters$Composition ~ painters$School, main="Boxplot of Composition by groups of school", xlab="painters$School", ylab = "painters$Composition", boxwex = .4)
``` 

4. Consider the following distribution with density function
$$
f(y) = \frac{r\beta^r}{(\beta+y)^{r+1}}, \quad y \geq 0.
$$
Use the inverse transform method to generate 1000 random observations from this distribution with $r = 3$ and $\beta = 4$. Print out the first 8 observations of the generated sample. Plot the density histogram of the total sample and plot the sample density curve on the histogram.
    
```{r indent=indent1, out.width="50%"}
set.seed(24)
# Put your code here
n <- 1000;
u <- runif(n);
x <- ((192/u)^0.25)-4;
# Print out the first 8 observations of the generated sample.
head(x, 8)
# Plot the density histogram of the total sample.
hist(x, prob = TRUE)
# Plot the sample density curve on the histogram.
lines(density(x))
``` 

5. We want to compute a Monte Carlo estimate of
$$
\int_{0}^{\pi / 6} x\cos (x) \, dx
$$
using the importance sampling method. We choose three importance functions $f_1 = 1$ which is the density of Uniform(0,1), $f_2 =6/\pi$ which is the density of Uniform(0, $\pi/6$) and $f_3 = 2\cos (x), 0 <x<\pi/6$. Show your estimation results and their corresponding standard errors using 10000 replications. **(Hint: the code of the inverse of $sin()$ is $asin()$)**
    
```{r indent=indent1}
set.seed(47)
# Put your code here
m <- 10000
theta.hat <- se <- numeric(3)
g <- function(x){
  x*cos(x)*(x>0)*(x<pi/6)
}

# here we use f1.
x <- runif(m)
i <- c(which(x > pi/6), which(x < 0)) #indicator function.
x[i] <- 2 #to catch overflow errors in g(x)
fg <- g(x)
theta.hat[1] <- mean(fg)
se[1] <- sd(fg)

# here we use f2.
x <- runif(m, 0, pi/6)
fg <- g(x) / (6/pi)
theta.hat[2] <- mean(fg)
se[2] <- sd(fg)

# here we use f3.
u <- runif(m) #inverse transform method
x <- asin(u/2)
#i <- c(which(x > pi/6), which(x < 0)) #indicator function.
#x[i] <- 2 #to catch overflow errors in g(x)
fg <- x/2
#fg <- g(x) / (2*cos(x))
theta.hat[3] <- mean(fg)
se[3] <- sd(fg)

rbind(theta.hat,se)
```       
    
    
